---
title: "NBdata Models"
author: "Marina Peñalver Ripoll"
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    code_folding: show
    theme: yeti
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(2512)
```


## Introduction

In this document, we will use the `maSigPro` package to analyze the evolution of one gene.


First, we load the necessary libraries for the analysis.

```{r load-libraries, include=FALSE}
library(maSigPro)
library(ggplot2)
library(patchwork)
library(splines)
library(mgcv)
library(MASS)
```


We load the dataset provided by the `maSigPro` package, which includes the expression data.

```{r load-data}
data(NBdata)
data(NBdesign)
```

Create the design matrix from the experimental design.

```{r design-matrix}
d <- make.design.matrix(NBdesign)
design <- as.data.frame(NBdesign)
```



We define functions to plot the expression of specific genes across time points for two groups.

```{r plot-gene-functions}
plot_gene <- function(gene_number, design, NBdata) {
  gene_data <- data.frame(
    Time = design$Time[1:18],
    Group1 = NBdata[gene_number, 1:18],
    Group2 = NBdata[gene_number, 19:36]
  )
  
  data_mean <- data.frame(
    Time = c(0, 12, 24, 36, 48, 60),
    Mean.G1 = sapply(split(gene_data$Group1, gene_data$Time), mean),
    Mean.G2 = sapply(split(gene_data$Group2, gene_data$Time), mean)
  )
  
  plot <- ggplot(gene_data) +
    geom_point(data = gene_data, aes(x = Time, y = Group1, color = "Group 1")) +
    geom_point(data = gene_data, aes(x = Time, y = Group2, color = "Group 2")) +
    geom_line(data = data_mean, aes(x = Time, y = Mean.G1), linewidth = 1, color = "coral") +
    geom_line(data = data_mean, aes(x = Time, y = Mean.G2), linewidth = 1, color = "steelblue2") +
    labs(title = paste("Gene", gene_number),
         x = "Time",
         y = "Expression",
         color = NULL) +
    scale_color_manual(values = c("Group 1" = "coral", "Group 2" = "steelblue2")) +
    theme_light()
}

plot_genes <- function(gene_numbers, design, NBdata, ncol, nrow) {
  plots <- list()
  j = 1
  for (i in gene_numbers) {
    plots[[j]] <- plot_gene(i, design, NBdata)
    j = j + 1
  }
  combined_plot <- wrap_plots(plots, ncol = ncol, nrow = nrow)
  return(combined_plot)
}
```


And we define some functions for plotting the models:
```{r}
plot_lm <- function(model, design=design, title=NULL){
  data <- data.frame(
    Group1 = model$fitted.values[1:18],
    Group2 = model$fitted.values[19:36],
    Time = design$Time[1:18]
  )
  
  n = 200
  time2 <- seq(from = 0, to = 60, length.out = n)
  data2 <- data.frame(
    time2 = time2,
    pred.G1 = predict(model, data.frame(Time = time2, Group=as.factor(rep("Group.1", n)))),
    pred.G2 = predict(model, data.frame(Time = time2, Group=as.factor(rep("Group.2", n))))
  )
  
  plot <- ggplot(data) +
    geom_point(data = gene_data, aes(x = Time, y = Group1, color = "Grupo 1"), size=2.5, shape=18) +
    geom_point(data = gene_data, aes(x = Time, y = Group2, color = "Grupo 2"), size=2.5, shape=18) +
    geom_point(aes(x = Time, y = Group1, color = "Grupo 1"), size=3) +
    geom_point(aes(x = Time, y = Group2, color = "Grupo 2"), size=3) +
    geom_line(data=data2, aes(x = time2, y = pred.G1), linewidth = 1, color="coral") +
    geom_line(data=data2, aes(x = time2, y = pred.G2), linewidth = 1, color="steelblue2") +
    labs(title = title,
         x = "Tiempo",
         y = "Expresión",
         color = NULL) +
    scale_color_manual(values = c("Grupo 1" = "coral", "Grupo 2" = "steelblue2")) +
    theme_light()
  
  return(plot)
}


plot_glm <- function(model, design=design, title=NULL){
  data <- data.frame(
    Group1 = model$fitted.values[1:18],
    Group2 = model$fitted.values[19:36],
    Time = design$Time[1:18]
  )
  
  n = 200
  time2 <- seq(from = 0, to = 60, length.out = n)
  data2 <- data.frame(
    time2 = time2,
    pred.G1 = predict(model, data.frame(Time = time2, Group=as.factor(rep("Group.1", n))), type="response"),
    pred.G2 = predict(model, data.frame(Time = time2, Group=as.factor(rep("Group.2", n))), type="response")
  )
  
  plot <- ggplot(data) +
    geom_point(data = gene_data, aes(x = Time, y = Group1, color = "Grupo 1"), size=2.5, shape=18) +
    geom_point(data = gene_data, aes(x = Time, y = Group2, color = "Grupo 2"), size=2.5, shape=18) +
    geom_point(aes(x = Time, y = Group1, color = "Grupo 1"), size=3) +
    geom_point(aes(x = Time, y = Group2, color = "Grupo 2"), size=3) +
    geom_line(data=data2, aes(x = time2, y = pred.G1), linewidth = 1, color="coral") +
    geom_line(data=data2, aes(x = time2, y = pred.G2), linewidth = 1, color="steelblue2") +
    labs(title = title,
         x = "Tiempo",
         y = "Expresión",
         color = NULL) +
    scale_color_manual(values = c("Grupo 1" = "coral", "Grupo 2" = "steelblue2")) +
    theme_light()
  
  return(plot)
}
```




# Linear and Polynomial Regression

We fit linear and polynomial regression models to the data and plot the results.

```{r linear-polynomial-regression}
# Select a gene for analysis
gen <- 13
y <- as.numeric(NBdata[gen,])
Time <- NBdesign[,1]
Group <- factor(NBdesign[, 3] + 2 * NBdesign[, 4], 
                labels = colnames(NBdesign[, 3:4]))

data <- data.frame(
  y = y,
  Time = Time,
  Group = Group
)

gene_data <- data.frame(
  Time = design$Time[1:18],
  Group1 = NBdata[gen, 1:18],
  Group2 = NBdata[gen, 19:36]
)
```



```{r linear-models}
# Linear regression models
lm1 <- lm(y ~ Time + Group)
lm2 <- lm(y ~ Time * Group)
# summary(lm1)
# summary(lm2)
anova(lm1, lm2)
```

```{r polynomial-models}
# Polynomial regression models
lm3 <- lm(y ~ poly(Time, 2) + Group)
lm4 <- lm(y ~ poly(Time, 2) * Group)
# summary(lm3)
# summary(lm4)
anova(lm3, lm4)
```


We can search the optimal degree for polynomial regression:
```{r}
max_degree <- 5
aic_values <- numeric(max_degree)
bic_values <- numeric(max_degree)
sce_values <- numeric(max_degree)

for (degree in 1:max_degree) {
  model <- lm(y ~ poly(Time, degree) * Group)
  aic_values[degree] <- AIC(model)
  bic_values[degree] <- BIC(model)
  sce_values[degree] <- sum((model$fitted.values - y)^2)
}

optimal_degree <- which.min(aic_values)
optimal_lm <- lm(y ~ poly(Time, optimal_degree) * Group)
print(plot_lm(optimal_lm, design = design, 
              title = paste("Polynomial Regression of degree", optimal_degree)))
```


# Generalized Linear Models (GLM)

We fit GLM models to the data and plot the results.

```{r glm-analysis}
glm1 <- glm(y ~ Time + Group, family = poisson)
glm2 <- glm(y ~ Time * Group, family = poisson)
# summary(glm1)
# summary(glm2)
anova(glm1, glm2, test = "Chisq")

# Polynomial GLM models
glm3 <- glm(y ~ poly(Time, 2) + Group, family = poisson)
glm4 <- glm(y ~ poly(Time, 2) * Group, family = poisson)
# summary(glm3)
# summary(glm4)
anova(glm3, glm4, test = "Chisq")

# Optimal degree for polynomial GLM
max_degree <- 5
aic_values <- numeric(max_degree)
bic_values <- numeric(max_degree)
sce_values <- numeric(max_degree)

for (degree in 1:max_degree) {
  model <- glm(y ~ poly(Time, degree) * Group, family = poisson)
  aic_values[degree] <- AIC(model)
  bic_values[degree] <- BIC(model)
  sce_values[degree] <- sum((model$fitted.values - y)^2)
}

optimal_degree <- which.min(aic_values)
optimal_glm <- glm(y ~ poly(Time, optimal_degree) * Group, family = poisson)
print(plot_glm(optimal_glm, design = design, 
               title = paste("Polynomial GLMof degree", optimal_degree)))
```


# Generalized Additive Models (GAM)

We fit GAM models to the data using natural and cubic splines, and plot the results.

```{r gam-analysis}
# Function to evaluate GAM models
evaluate_gam <- function(y, Time, Group, knot_range, family = poisson(), type = "natural") {
  aic_values <- numeric(length(knot_range))
  bic_values <- numeric(length(knot_range))
  models <- vector("list", length(knot_range))
  sce <- numeric(length(knot_range))
  
  for (i in seq_along(knot_range)) {
    k <- knot_range[i]
    if (type == "natural") {
      gam_model <- gam(y ~ ns(Time, df = k + 1) * Group, family = family, 
                       data = data.frame(Time = Time, Group = Group, y = y)) }
    else if (type == "cubic") {
      gam_model <- gam(y ~ bs(Time, df = k + 3) * Group, family = family, 
                       data = data.frame(Time = Time, Group = Group, y = y)) }
    aic_values[i] <- AIC(gam_model)
    bic_values[i] <- BIC(gam_model)
    models[[i]] <- gam_model
    sce[i] <- sum((gam_model$fitted.values - y)^2)
  }
  
  optimal_index <- which.min(aic_values)
  optimal_knots <- knot_range[optimal_index]
  optimal_model <- models[[optimal_index]]
  
  return(list(optimal_knots = optimal_knots, optimal_model = optimal_model, aic_values = aic_values, sce = sce))
}

knot_range = 1:4
```

## Cubic spline

We evaluate GAM models with cubic spline:

```{r}
evaluate_gam(y, Time, Group, knot_range, type = "cubic")
```

And we fit the model with 2 knots
```{r}
optimal_cubic_gam <- gam(y ~ bs(Time, df = 2 + 3) * Group, family = poisson, data = data)
```

We can visualize the model:
```{r}
n <- 200
time2 <- seq(from = 0, to = 60, length.out = n)
data_bs1 <- data.frame(Time = time2, Group = as.factor(rep("Group.1", n)))
data_bs1 <- cbind(data_bs1, bs(time2, df = 2 + 3))
data_bs2 <- data.frame(Time = time2, Group = as.factor(rep("Group.2", n)))
data_bs2 <- cbind(data_bs2, bs(time2, df = 2 + 3))

data1 <- data.frame(
  Group1 = optimal_cubic_gam$fitted.values[1:18],
  Group2 = optimal_cubic_gam$fitted.values[19:36],
  Time = design$Time[1:18])

data2 <- data.frame(
  time2 = time2,
  pred.G1 = predict(optimal_cubic_gam, newdata = data_bs1, type = "response"),
  pred.G2 = predict(optimal_cubic_gam, newdata = data_bs2, type = "response"))

ggplot(data1) +
  geom_point(data = gene_data, aes(x = Time, y = Group1, color = "Group 1"), size = 2.5, shape = 18) +
  geom_point(data = gene_data, aes(x = Time, y = Group2, color = "Group 2"), size = 2.5, shape = 18) +
  geom_point(aes(x = Time, y = Group1, color = "Group 1"), size = 2.5) +
  geom_point(aes(x = Time, y = Group2, color = "Group 2"), size = 2.5) +
  geom_line(data = data2, aes(x = time2, y = pred.G1), linewidth = 1, color = "coral") +
  geom_line(data = data2, aes(x = time2, y = pred.G2), linewidth = 1, color = "steelblue2") +
  labs(title = "GAM with Cubic Spline",
       x = "Time",
       y = "Expression",
       color = NULL) +
  scale_color_manual(values = c("Group 1" = "coral", "Group 2" = "steelblue2")) +
  theme_light()
```



## Natural cubic splines

Now, we will use natural cubic spline in our model
```{r}
evaluate_gam(y, Time, Group, knot_range, type = "natural")
```

The optimal model has 4 knots:
```{r}
optimal_natural_gam <- gam(y ~ ns(Time, df = 4+1) * Group, family = poisson, data = data)
```

The graph is:
```{r}
n <- 200
time2 <- seq(from = 0, to = 60, length.out = n)
data_ns1 <- data.frame(Time = time2, Group = as.factor(rep("Group.1", n)))
data_ns1 <- cbind(data_ns1, ns(time2, df = 4 + 1))
data_ns2 <- data.frame(Time = time2, Group = as.factor(rep("Group.2", n)))
data_ns2 <- cbind(data_ns2, ns(time2, df = 4 + 1))

data1 <- data.frame(
  Group1 = optimal_natural_gam$fitted.values[1:18],
  Group2 = optimal_natural_gam$fitted.values[19:36],
  Time = design$Time[1:18])

data2 <- data.frame(
  time2 = time2,
  pred.G1 = predict(optimal_natural_gam, newdata = data_ns1, type

 = "response"),
  pred.G2 = predict(optimal_natural_gam, newdata = data_ns2, type = "response"))

ggplot(data1) +
  geom_point(data = gene_data, aes(x = Time, y = Group1, color = "Group 1"), size = 2.5, shape = 18) +
  geom_point(data = gene_data, aes(x = Time, y = Group2, color = "Group 2"), size = 2.5, shape = 18) +
  geom_point(aes(x = Time, y = Group1, color = "Group 1"), size = 2.5) +
  geom_point(aes(x = Time, y = Group2, color = "Group 2"), size = 2.5) +
  geom_line(data = data2, aes(x = time2, y = pred.G1), linewidth = 1, color = "coral") +
  geom_line(data = data2, aes(x = time2, y = pred.G2), linewidth = 1, color = "steelblue2") +
  labs(title = "GAM with Natural Spline",
       x = "Time",
       y = "Expression",
       color = NULL) +
  scale_color_manual(values = c("Group 1" = "coral", "Group 2" = "steelblue2")) +
  theme_light()
```