---
title: "NBdata"
author: "Marina Peñalver Ripoll"
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    code_folding: show
    theme: yeti
---


# Introduction

In this document, we will use the `maSigPro` package to analyze time course data and identify significant genes.


First, we load the necessary libraries for the analysis.

```{r load-libraries, include=FALSE}
library(maSigPro)
library(ggplot2)
library(patchwork)
```


We load the dataset provided by the `maSigPro` package, which includes the expression data (`NBdata`) and the experimental design (`NBdesign`).

```{r load-data}
data(NBdata)
data(NBdesign)
```


This dataset is part of a larger simulated and normalized dataset with 2 experimental groups, 6 timepoints and 3 replicates. Simulation has been done by using a negative binomial distribution. The first 20 genes are simulated with changes among time. Preview the first few rows of the data and the design matrix:

```{r preview-data}
head(NBdata)
head(NBdesign)
```


We create a design matrix from the experimental design.

```{r design-matrix}
d <- make.design.matrix(NBdesign)
design <- as.data.frame(NBdesign)
```

# Plotting Gene Expression

We define a function to plot the expression of a specific gene across time points for two groups.

```{r plot-gene, fig.align='center', fig.width=6, fig.height=4}
plot_gene <- function(gene_number, design, NBdata) {
  
  gene_data <- data.frame(
    Time = design$Time[1:18],
    Group1 = NBdata[gene_number, 1:18],
    Group2 = NBdata[gene_number, 19:36]
  )
  
  data_mean <- data.frame(
    Time = c(0, 12, 24, 36, 48, 60),
    Mean.G1 = sapply(split(gene_data$Group1, gene_data$Time), mean),
    Mean.G2 = sapply(split(gene_data$Group2, gene_data$Time), mean)
  )
  
  plot <- ggplot(gene_data) +
    geom_point(data=gene_data, aes(x = Time, y = Group1, color = "Group 1")) +
    geom_point(data=gene_data, aes(x = Time, y = Group2, color = "Group 2")) +
    geom_line(data=data_mean, aes(x = Time, y = Mean.G1), linewidth = 1, 
              color="coral") +
    geom_line(data=data_mean, aes(x = Time, y = Mean.G2), linewidth = 1, 
              color="steelblue2") +
    labs(title = paste("Gene", gene_number),
         x = "Time",
         y = "Expression",
         color = NULL) +
    scale_color_manual(values = c("Group 1" = "coral", 
                                  "Group 2" = "steelblue2")) +
    theme_light()
}

# Plot a specific gene
plot <- plot_gene(52, design, NBdata)
print(plot)
```


We define a function to combine plots of multiple genes into a single plot layout.

```{r combine-plots, fig.align='center', fig.width=10, fig.height=5}
plot_genes <- function(gene_numbers, design, NBdata, ncol, nrow) {
  plots <- list()
  j = 1
  for (i in gene_numbers) {
    plots[[j]] <- plot_gene(i, design, NBdata)
    j = j + 1
  }
  combined_plot <- wrap_plots(plots, ncol=ncol, nrow=nrow)
  return(combined_plot)
}

# Combine plots of selected genes
combined_plot <- plot_genes(c(1:3, 11:13), design, NBdata, ncol = 3, nrow=2)
print(combined_plot)
```

# Selecting Significant Genes

We perform regression fitting for each gene and identify significant genes.

* The `p.vector` function performs a regression fit for each gene, taking all variables present in the model given by a regression matrix, and returns a list of False Discovery Rate (FDR) corrected significant genes.

    ```{r significant-genes}
    fit <- p.vector(NBdata, d, Q = 0.05, MT.adjust = "BH", min.obs = 20, 
                    family=poisson())
    rownames(fit$SELEC)
    ```


* The `T.fit` function performs stepwise regression to refine the model, using methods like backward elimination, and returns a list of significant gene profiles.

    ```{r stepwise-regression}
    tstep <- T.fit(fit, step.method = "backward", alfa = 0.05, family=poisson())
    rownames(tstep$sig.profiles)
    ```


* The `get.siggenes` function extracts the significant genes based on the $R^2$ value.

    ```{r extract-significant-genes}
    sigs <- get.siggenes(tstep, rsq = 0.7, vars = "groups")
    sigs$summary
    ```
    
    The value of the $R^2$ for each gene is:
    ```{r R-square}
    tstep$sol$`R-squared`
    ```



The significant genes are:
```{r plot-significant-genes, fig.align='center', fig.width=15, fig.height=10}
combined_plot <- plot_genes(c(1:14, 18, 19), design, NBdata, ncol = 4, nrow=4)
print(combined_plot)
```

And the gene that we exclude are:
```{r plot-nonsignificant-genes, fig.align='center', fig.width=10, fig.height=5}
combined_plot <- plot_genes(c(15, 16, 17, 20), design, NBdata, ncol = 2, nrow=2)
print(combined_plot)
```



Finally, we select a gene for further study:
```{r new-seed, include=FALSE}
set.seed(2512)
```

```{r random-gene, fig.align='center', fig.width=8, fig.height=4}
gen <- sample(c(1:14, 18, 19), 1)
plot <- plot_gene(gen, design, NBdata)
print(plot)
```





